---
title: "EDS 222: Homework 1"
date: "Assigned ???, due ???"
author: "Your Name Here"
editor_options: 
  chunk_output_type: console
---

## Background

*(The case study in this exercise is based on reality, but does not include actual observational data.)*

In this exercise we will look at a case study concerning air quality in South Asia. The World Health Organization estimates that air pollution kills an estimated seven million people per year, due to its effects on the cardiovascular and respiratory systems. Out of the 40 most polluted cities in the world, South Asia is home to 37, and Pakistan was ranked to contain the second most air pollution in the world in 2020 (IQAIR, 2020). In 2019, Lahore, Pakistan was the 12th most polluted city in the world, exposing a population of 11.1 million people to increased mortality and morbidity risks.

In this exercise, you are given two datasets from Lahore, Pakistan and are asked to compare the two different data collection strategies from this city. These data are:

-   Crowd-sourced data from air quality monitors located in people's homes. These data are voluntarily collected by individual households who choose to install a monitor in their home and upload their data for public access.

-   Official government data from monitors installed by government officials at selected locations across Lahore. There have been reports that government officials strategically locate monitors in locations with cleaner air in order to mitigate domestic and international pressure to clean up the air.

::: callout-note
All data for EDS 222 will be stored on the Taylor server, in the shared `/courses/eds-222/data/` directory. Please see material from EDS 214 on how to access and retrieve data from Taylor. These data are small; all compute can be handled locally. Thanks to Bren PhD student Fatiq Nadeem for assembling these data!
:::

In answering the following questions, please consider the lecture content from class on sampling strategies, as well as the material in Chapter 2 of [*Introduction to Modern Statistics*](https://openintro-ims.netlify.app/data-design). Include in your submission your version of this file "`eds-222-hw1.qmd`" and the rendered HTML output, each containing complete answers to all questions *as well as the associated code*. Questions with answers unsupported by the code will be marked incomplete. Showing your work this way will help you develop the habit of creating reproducible code.

## Assessment

#### Load Packages
```{r}
library(dplyr)
library(ggplot2)
library(sf)
```

### Question 1

Load the data from each source and label it as `crowdsourced` and `govt` accordingly. For example:

```{r}
crowdsourced <- readRDS(file.path("data", "HW1", "airpol-PK-crowdsourced.RDS"))
govt <- readRDS(file.path("data", "HW1","airpol-PK-govt.RDS"))
```

::: callout-warning
There's an implicit assumption about file organization in the code above. What is it? How can you make the code work?
:::

**Answer:** The code assumes that we are getting files from our `../data/` folder. I adjusted it to add the `../data/HW1/` sub-folder that I pulled from Cyberduck with all the important data. The file type this read function expects is an `.RDS` type. 

1.  These dataframes have one row per pollution observation. How many pollution records are in each dataset?

**Answer:** 

```{r}
# Using nrow() to check the row count of each dataframe.
cat("The `crowdsourced` dataframe has", nrow(crowdsourced), "pollution records.\n")
cat("The `govt` dataframe has", nrow(govt), "pollution records.")
```

2.  Each monitor is located at a unique latitude and longitude location. How many unique monitors are in each dataset?

::: callout-tip
`group_by(longitude,latitude)` and `cur_group_id()` in `dplyr` will help in creating a unique identifier for each (longitude, latitude) pair.
:::

**Answer:**

```{r}
# Group our dataframes by group_by(long, lat) 
# mutate() to add an `monitor_id` column to differentiate each monitor by grouping.
crowdsourced_unique_monitors <- crowdsourced %>%
  group_by(longitude, latitude) %>%
  mutate(monitor_id = cur_group_id())

# Doing the same thing to this dataframe. 
govt_unique_monitors <- govt %>%
  group_by(longitude, latitude) %>%
    mutate(monitor_id = cur_group_id())

# Printing out the number of unique monitors by using n_distinct() to sum 
# unique values in our newly created `monitor_id` column.
cat("The `crowdsourced` dataframe has", n_distinct(crowdsourced_unique_monitors$monitor_id), "unique monitors.\n")
cat("The `govt` dataframe has", n_distinct(govt_unique_monitors$monitor_id), "unique monitors.")
```

### Question 2

The goal of pollution monitoring in Lahore is to measure the average pollution conditions across the city.

1.  What is the *population* in this setting? Please be precise.

**Answer:** The population we are taking a look at is the entirety of the 
air quality data in Lahore, Pakistan from Nov 4th 2018 to Nov 30th 2019. 

2.  What are the *samples* in this setting? Please be precise.

**Answer:** Our population is distributed into to samples: 

- `Crowdsourced monitors:` Which consist of monitors voluntarily installed by individuals which reflects air quality in homes across and localized areas of 
Lahore.

- `Government monitors:` Which consist of monitors installed at specific outdoor
locations by government officials. We plan to question whether these monitors 
could be strategically placed in areas with relatively cleaner air.

3.  These samples were not randomly collected from across locations in Lahore. Given the sampling approaches described above, discuss possible biases that may enter when we use these samples to construct estimates of population parameters.

**Answer:** Given the sampling techniques described we can potentially perceive
a selection bias in the `crowdsourced` data where people who can afford to 
purchase monitors may cause wealthier or environmentally conscious populations 
to become over represented. Another bias we can potentially perceive is a location
bias in both our data-sets where our `crowdsourced` indoor air quality 
measurements may not accurately reflect outdoor conditions. Furthermore, in 
reference to the `government` data, the officials could have placed these 
monitors in strategic locations where the air quality is cleaner.

### Question 3

1.  For both the government data and the crowd-sourced data, report the sample mean, sample minimum, and sample maximum value of PM 2.5 (measured in $\mu g/m^3$).

**Answer:**

```{r}
# Using the summarize() function to find the mean, min, and max PM values.
crowdsourced_summary <- crowdsourced_unique_monitors %>%
  ungroup() %>% # Remove our previous (long, lat) grouping
  summarize(PM_mean = mean(PM),
            PM_min = min(PM),
            PM_max = max(PM))

# Print crowdsourced mean, min, and max
cat("Crowdsourced data", "PM Mean:", crowdsourced_summary$PM_mean, "|",
  "PM Min:", crowdsourced_summary$PM_min, "|",
  "PM Max:", crowdsourced_summary$PM_max, "\nMeasured in \\mu g/m^3", "\n")

# Same process as above 
govt_summary <- govt_unique_monitors %>%
  ungroup() %>% # Remove our previous (long, lat) grouping
  summarize(PM_mean = mean(PM),
            PM_min = min(PM),
            PM_max = max(PM))

# Print government mean, min, and max
cat("Government data", "PM Mean:", govt_summary$PM_mean, "|",
  "PM Min:", govt_summary$PM_min, "|",
  "PM Max:", govt_summary$PM_max, "\nMeasured in \\mu g/m^3", "\n")
```

2.  Discuss any key differences that you see between these two samples.

**Answer:** The `government` data represents a **\~56.58%** decrease in mean 
2.5PM values. Noticeably the max for `government` data is less than the mean 
for the `crowdsourced` data. The `crowdsourced` maximum is almost double the government max. The mins for both data-sets are similar.

3.  Are the differences in mean pollution as expected, given what we know about the sampling strategies?

**Answer:** Yes, the mean particulate matter from our `government` collected 
PM2.5 is **\~56.28%** less at an average of **\~29.65** $\mu g/m^3$ which is 
a pretty drastic difference indicating that there is indeed a strategic role 
played by the government when collecting air quality data in Lahore.

### Question 4

Use the location of the air pollution stations for both of the sampling strategies to generate a map showing locations of each observation. Color the two samples with different colors to highlight how each sample obtains measurements from different parts of the city.

::: callout-tip
`longitude` indicates location in the *x*-direction, while `latitude` indicates location in the *y*-direction. With `ggplot2` this should be nothing fancy. We'll do more spatial data in `R` later in the course.
:::

```{r}
merged_df <- bind_rows(
  (crowdsourced_unique_monitors %>%
     ungroup() %>%
     mutate(source = "crowd")),
  (govt_unique_monitors %>%
     ungroup() %>%
     mutate(source = "govt")))
```

```{r}
# Turned our df into an sf object with longitude and latitude as geometry
merged_sf <- st_as_sf(merged_df, coords = c("longitude", "latitude"), crs = 4326)

# Plot
ggplot() +
  geom_sf(data = merged_sf, aes(size = PM, color = source), alpha = 0.5) +
  labs(title = "Lahore's PM data in micro grams per cubic meter ",
    x = "Longitude", 
    y = "Latitude") +
  theme_bw()
```

### Question 5

The local newspaper in Pakistan, *Dawn*, claims that the government is misreporting the air pollution levels in Lahore. Do the locations of monitors in question 4, relative to crowd-sourced monitors, suggest anything about a possible political bias?

**Answer: ** Based on the plot in Question 4 if the government stations are concentrated in areas with less pollution (which they are) this would suggest that the government is selectively placing monitors to under report pollution levels. The crowd-sourced monitors are spread distributed pretty evenly which could indicate a more accurate representation of air quality. We can better see that the government is using a selection/location bias.

### Question 6

Given the recent corruption in air quality reporting, the Prime Minister of Pakistan has hired an independent body of environmental data scientists to create an unbiased estimate of the mean PM 2.5 across Lahore using some combination of both government stations and crowd sourced observations.

NASA's satellite data indicates that the average PM across Lahore is 89.2 $\mu g/m^3$. Since this is the most objective estimate of population-level PM 2.5 available, your goal is to match this mean as closely as possible by creating a new ground-level monitoring sample that draws on both the government and crowd-sourced samples.

#### Question 6.1

First, generate a *random sample* of size $n=1000$ air pollution records by (i) pooling observations across the government and the crowd-sourced data; and (ii) drawing observations at random from this pooled sample.

::: callout-tip
`bind_rows()` may be helpful.
:::

```{r}
# Setting the seed for our random sample
set.seed(1)

# Generate a df with a random sample of n = 1000
random <- merged_df %>%
  sample_n(size = 1000, replace = TRUE)

# Print mean PM of sample
cat("Mean PM of n = 1000 sample:", mean(random$PM), "\\mu g/m^3")
```

Second, create a *stratified random sample*. Do so by (i) stratifying your pooled data-set into strata of 0.01 degrees of latitude, and (ii) randomly sampling 200 air pollution observations from each stratum.

```{r}
merged_df <- merged_df %>%
  mutate(stratum = floor(latitude * 100) / 100)

strat_df <- merged_df %>%
  group_by(stratum) %>%
  sample_n(size = 200, replace = TRUE) %>%
  ungroup()

# Print mean PM of sample
cat("Mean PM of stratified n = 200 sample:", mean(strat_df$PM), "\\mu g/m^3")
```

#### Question 6.2

Compare estimated means of PM 2.5 for each sampling strategy to the NASA estimate of 89.2 $\mu g/m^3$. Which sample seems to match the satellite data best? What would you recommend the Prime Minister do? Does your proposed sampling strategy rely more on government or on crowd-sourced data? Why might that be the case?


**Answer:** Our stratified sample mean **64.125 µg/m³** is closer to the NASA estimate. This suggests that location-based sampling is more effective in representing the pollution level The PM across Lahore. The Prime Minister should start using stratified sampling to ensure a more reliable estimate of air quality.

